\MakeUppercase{Graph Product Structure for Non-Minor-Closed Classes}

### 2022-03-18 k-planar.tex
added comments about the framed proof

The starting point for this work is the following `product structure theorem' for planar graphs\footnote{In this paper, all graphs are finite and undirected. Unless mentioned otherwise, all graphs are also simple. For any graph $G$ and any set $S$ (typically $S\subseteq V(G)$), let $G[S]$  denote the graph with vertex set $V(G)\cap S$ and edge set $\{uv\in E(G) : u,v\in S\}$.  We use $G-S$ as a shorthand for $G[V(G)\setminus S]$. Undefined terms are in \citep{Diestel5}.} by \citet{DJMMUW20} (with an improvement by \citet{UWY}). A graph $G$ is \defin{contained in} a graph $X$ if $G$ is isomorphic to a subgraph of $X$.

@@

\noindent\cref{PlanarProduct} has been the key tool in solving the following well-known open problems:
\begin{compactitem}
\item \citet{DJMMUW20} use it to prove that planar graphs have bounded queue-number (resolving a conjecture of \citet{HLR92}).
\item  \citet{dujmovic.esperet.ea:planar} use it to prove that planar graphs have bounded non-repetitive chromatic number (resolving a conjecture of \citet{AGHR-RSA02}).
\item \citet{DEJGMM21} have since used it to find asymptotically optimal adjacency labellings of planar graphs (resolving a problem of \cite{kannan.naor.ea:implicit}).
\item \citet{EJM} use it to show the existence of a 'universal graph' with $n^{1+o(1)}$ vertices and edges that contains every $n$-vertex planar graph as an induced subgraph (resolving a problem of \citet{babai.chung.ea:on}).
\end{compactitem}
In addition to these rather well-known problems, \cref{PlanarProduct} has been used to resolve or make substantial progress on a number of other problems on planar graphs, including $p$-centered colouring \cite{DFMS21}, $\ell$-vertex ranking~\citep{BDJM}, twin-width~\citep{BKW}, and odd colourings~\citep{DMO}.

@@

All of these results hold for any graph class that has a product structure theorem analogous to \cref{PlanarProduct}; that is, for any graph class  $\mathcal{G}$ where every graph in $\mathcal{G}$ is contained in $H\boxtimes P\boxtimes K_\ell$ where $H$ has bounded treewidth, $P$ is a path, and $\ell$ is bounded.\footnote{It is easily seen that $\tw(H\boxtimes K_\ell) \leq (\tw(H)+1)\ell-1$, so we may assume that $\ell=1$ in this definition.} These applications motivate finding product structure theorems for other graph classes. \citet{DJMMUW20} prove product structure theorems for graphs of bounded Euler genus\footnote{The \textit{Euler genus} of the orientable surface with $h$ handles is $2h$. The \textit{Euler genus} of the non-orientable surface with $c$ cross-caps is $c$. The \textit{Euler genus} of a graph $G$ is the minimum integer $g$ such that $G$ embeds in a surface of Euler genus $g$. Of course, a graph is planar if and only if it has Euler genus 0; see \citep{mohar.thomassen:graphs} for more about graph embeddings in surfaces.} and for apex-minor-free graphs\footnote{A graph $M$ is a \textit{minor} of a graph $G$ if a graph isomorphic to $M$ can be obtained from a subgraph of $G$ by contracting edges. A class $\mathcal{G}$ of graphs is \defin{minor-closed} if for every graph $G\in\mathcal{G}$, every minor of $G$ is in $\mathcal{G}$. A minor-closed class is \defin{proper} if it is not the class of all graphs. For example, for fixed $g\geq 0$, the class of graphs with Euler genus at most $g$ is a proper minor-closed class. A graph $G$ is $t$-apex if it contains a set $A$ of at most $t$ vertices such that $G-A$ is planar. A 1-apex graph is \defin{apex}.  A minor-closed class $\mathcal{G}$ is apex-minor-free if some apex graph is not in $\mathcal{G}$.}, and \citet{DEMWW22} do so for graphs in any minor-closed class and with bounded maximum degree. See \cref{Generalisations} for more precise statements and see \citep{DHJLW21} for a survey on this topic.

@@

This definition is related to $k$-planarity because of the following observation:

@@

A \defin{layering} of a graph $G$ is a sequence $\mathcal{L}=(L_0,L_1,\ldots)$ such that $\{L_0,L_1,\ldots\}$ is a partition of $V(G)$ and for every edge $vw\in E(G)$, if $v\in L_i$ and $w\in L_j$ then $|j-i|\leq 1$.  For any partition $\PP=\{S_1,\ldots,S_p\}$ of $V(G)$, a \defin{quotient graph} $H=G/\PP$ has a $p$-element vertex set $V(H)=\{x_1,\ldots,x_p\}$ and $x_ix_j\in E(H)$ if and only if there exists an edge $vw\in E(G)$ such that $v\in S_i$ and $w\in S_j$. To highlight the importance of the quotient graph $H$, we call $\PP$ an \defin{$H$-partition} and write this concisely as $\PP=\{S_x : x\in V(H)\}$ so that each element of $\PP$ is indexed by the vertex it creates in $H$.

@@

The purpose of this section is to prove our main technical result, \cref{ShortcutPartition}. This theorem shows how, given a $(k,d)$-shortcut system $\SS$ of a graph $G$, an $H$-partition of $G$ can be used to obtain a $J$-partition of $G^{\SS}$ where the layered width  does not increase dramatically and the treewidth of $J$ is not much more than the treewidth of $H$.

@@

Consider a tree $T$ rooted at some node $x_0\in V(T)$. Each subtree $T'$ of $T$ is considered to be rooted at the node in $T'$ closest to $x_0$. A node $a\in V(T)$ is a \defin{$T$-ancestor} of $x\in V(T)$ (and $x$ is a \defin{$T$-descendant} of $a$) if $a$ is a vertex of the path, in $T$, from $x_0$ to $x$.  Note that each node $x\in V(T)$ is a $T$-ancestor and $T$-descendant of itself.  We say that a $T$-ancestor $a\in V(T)$ of $x\in V(T)$ is a \defin{strict} $T$-ancestor of $x$ if $a\neq x$.
The \defin{$T$-depth} of a node $x\in V(T)$ is the length of the path, in $T$, from $x_0$ to $x$.  For each node $x\in V(T)$, define
\[T_x := T[\{y\in V(T):\mbox{$x$ is a $T$-ancestor of $y$}\}] \]
to be the maximal subtree of $T$ rooted at $x$.

% \note{PM}{I have significantly simplified this section, which should make the referees happy. I will read it very carefully again tomorrow}

We begin with a standard technique that allows us to work with a normalised tree-decomposition whose tree has the same vertex set as the graph it decomposes:

@@

  Next, remove the node $x_0$ from $T_0$, and consider any node $a\in V(T_0)$ such that there is no vertex $x\in V(H)$ with $f(x)=a$.  In this case, $B_{a}\subseteq B_{a'}$ where $a'$ is the parent of $a$ since any $x\in B_a\setminus B_{a'}$ would have $f(x)=a$.  In this case, contract the edge $aa'$ in $T_0$, eliminating the node $a$.  Repeating this operation will eventually produce a width-$t$ tree-decomposition of $(B_x:x\in V(T_0))$ where $f$ is a bijection between $V(H)$ and $V(T_0)$.  Renaming each node $a\in V(T_0)$ as $f^{-1}(a)$ gives a tree-decomposition $(B_x:x\in V(T))$ with $V(T)=V(H)$.  By the definition of $f$, the tree-decomposition $(B_x:x\in V(T))$ satisfies \tref{subtree-root}.
\end{proof}

@@

@@

\begin{proof}
Let $\mathcal{L}:=\langle L_1,\ldots,L_h\rangle$ be a layering of $G$; let $\mathcal{Y}:=(Y_x: x\in V(H))$ be an $H$-partition of $G$ of layered width at most $\ell$ with respect to $\mathcal{L}$; and let $\mathcal{T}:=(B_x:x\in V(T))$ be a tree-decomposition of $H$ satisfying the conditions of \cref{nice-decomposition}.

For a node $x\in V(T)$, we say that a shortcut $P\in\SS$ \defin{crosses} $x$ if $Y_x$ contains an internal vertex of $P$.  In other words, if $P=v_0,\ldots,v_r$ and $\{v_1,\ldots,v_{r-1}\}\cap Y_x\neq\emptyset$.  We say that a vertex $v\in V(G)$ \defin{participates} in $x$ if $v\in Y_x$ or if $\SS$ contains a shortcut $P$ with $v\in V(P)$ and $P$ crosses $x$.  For each $v\in V(G)$, let $a(v)$ denote the lowest common $T$-ancestor of all the nodes $x\in V(T)$ in which $v$ participates.

For each $x\in V(T)$, define $S_x := \{v\in V(G): a(v)= x\}$. Since $a:V(G)\to V(T)$ is a function with domain $V(G)$ and range $V(T)$, $\PP:=(S_x : x\in V(T))$ is a sequence of pairwise disjoint sets that cover $V(G)$. Let $J:=G^\SS/\PP$ denote the resulting quotient graph. We consider $V(J)\subseteq V(T)$, where each $x\in V(J)$ is the vertex obtained by contracting $S_x$ in $G^{\SS}$. (Any node $x\in V(T)$ with $S_x=\emptyset$ does not contribute a vertex to $J$.)

From this point onward, the plan is to show that:
\begin{compactenum}[(i)]
  \item $\PP$ has small layered width with respect to the layering $\mathcal{L}$ of $G$ and that
  \item $J$ has small treewidth.
\end{compactenum}
Once we have established (i) and (ii), the result follows easily since a layering of $G^\SS$ is easily obtained from $\mathcal{L}$ by `compressing' groups of $k$ consecutive layers.  We begin with Step~(i):

>>>>

The starting point for this work is the following `product structure theorem' for planar graphs\footnote{In this paper, all graphs are finite and undirected. Unless mentioned otherwise, all graphs are also simple. For any graph $G$ and any set $S$ (typically $S\subseteq V(G)$), let $G[S]$  denote the graph with vertex set $V(G)\cap S$ and edge set $\{uv\in E(G) : u,v\in S\}$.  We use $G-S$ as a shorthand for $G[V(G)\setminus S]$. Undefined terms are in \citep{Diestel5}.} by \citet{DJMMUW20} (with an improvement by \citet{UWY}). A graph $G$ is \defin{contained} in a graph $X$ if $G$ is isomorphic to a subgraph of $X$.

@@

\noindent\cref{PlanarProduct} has been the key tool in solving the following well-known open problems:
\begin{compactitem}
\item \citet{DJMMUW20} use it to prove that planar graphs have bounded queue-number (resolving a conjecture of \citet{HLR92}).
\item  \citet{dujmovic.esperet.ea:planar} use it to prove that planar graphs have bounded non-repetitive chromatic number (resolving a conjecture of \citet{AGHR-RSA02}).
\item \citet{DEJGMM21} use it to find asymptotically optimal adjacency labellings of planar graphs (resolving a problem of \citet{kannan.naor.ea:implicit}).
\item \citet{EJM} use it to show the existence of a 'universal graph' with $n^{1+o(1)}$ vertices and edges that contains every $n$-vertex planar graph as an induced subgraph (resolving a problem of \citet{babai.chung.ea:on}).
\end{compactitem}
In addition, \cref{PlanarProduct} has been used to resolve or make substantial progress on a number of other problems on planar graphs, including $p$-centered colouring \cite{DFMS21}, $\ell$-vertex ranking~\citep{BDJM}, twin-width~\citep{BKW}, and odd colourings~\citep{DMO}.

@@

All of these results hold for any graph class that has a product structure theorem analogous to \cref{PlanarProduct}; that is, for any graph class  $\mathcal{G}$ where every graph in $\mathcal{G}$ is contained in $H\boxtimes P\boxtimes K_\ell$ where $H$ has bounded treewidth, $P$ is a path, and $\ell$ is bounded.\footnote{It is easily seen that $\tw(H\boxtimes K_\ell) \leq (\tw(H)+1)\ell-1$, so we may assume that $\ell=1$ in this statement.} These applications motivate finding product structure theorems for other graph classes. \citet{DJMMUW20} prove product structure theorems for graphs of bounded Euler genus\footnote{The \textit{Euler genus} of a surface with $h$ handles and $c$ crosscaps is $2h+c$. The \textit{Euler genus} of a graph $G$ is the minimum integer $g$ such that $G$ embeds in a surface of Euler genus $g$. Of course, a graph is planar if and only if it has Euler genus 0; see \citep{mohar.thomassen:graphs} for more about graph embeddings in surfaces.} and for apex-minor-free graphs\footnote{A graph $M$ is a \textit{minor} of a graph $G$ if a graph isomorphic to $M$ can be obtained from a subgraph of $G$ by contracting edges. A class $\mathcal{G}$ of graphs is \defin{minor-closed} if for every graph $G\in\mathcal{G}$, every minor of $G$ is in $\mathcal{G}$. A minor-closed class is \defin{proper} if it is not the class of all graphs. For example, for fixed $g\geq 0$, the class of graphs with Euler genus at most $g$ is a proper minor-closed class. A graph $G$ is $t$-apex if it contains a set $A$ of at most $t$ vertices such that $G-A$ is planar. A 1-apex graph is \defin{apex}.  A minor-closed class $\mathcal{G}$ is apex-minor-free if some apex graph is not in $\mathcal{G}$.}, and \citet{DEMWW22} do so for graphs in any minor-closed class and with bounded maximum degree. See \cref{Generalisations} for more precise statements and see \citep{DHJLW21} for a survey on this topic.

@@

This definition is related to $k$-planarity by the following observation:

@@

A \defin{layering} of a graph $G$ is a sequence $\mathcal{L}=\langle L_0,L_1,\ldots\rangle $ such that $\{L_0,L_1,\ldots\}$ is a partition of $V(G)$ 
\note{DW}{just write ``is an ordered partition $\mathcal{L}=\langle L_0,L_1,\ldots\rangle $ such that for every edge ...''}
and for every edge $vw\in E(G)$, if $v\in L_i$ and $w\in L_j$ then $|j-i|\leq 1$.  For any partition $\PP=\{S_1,\ldots,S_p\}$ of $V(G)$, a \defin{quotient graph} $H=G/\PP$ has a $p$-element vertex set $V(H)=\{x_1,\ldots,x_p\}$ and $x_ix_j\in E(H)$ if and only if there exists an edge $vw\in E(G)$ such that $v\in S_i$ and $w\in S_j$. To highlight the importance of the quotient graph $H$, we call $\PP$ an \defin{$H$-partition} and write this concisely as $\PP=\{S_x : x\in V(H)\}$ so that each element of $\PP$ is indexed by the vertex it creates in $H$ \note{DW}{``is indexed by the corresponding vertex in $H$''.}

@@

The purpose of this section is to prove our main technical result, \cref{ShortcutPartition}. This theorem shows how, given a $(k,d)$-shortcut system $\SS$ of a graph $G$, a $H$-partition of $G$ can be used to obtain a $J$-partition of $G^{\SS}$ where the layered width  does not increase dramatically and the treewidth of $J$ is not much more than the treewidth of $H$.

@@

Consider a tree $T$ rooted at some node $x_0\in V(T)$. Each subtree $T'$ of $T$ is considered to be rooted at the node in $T'$ closest to $x_0$. A node $a\in V(T)$ is a \defin{$T$-ancestor} of $x\in V(T)$ and $x$ is a \defin{$T$-descendant} of $a$ if $a$ is a vertex of the path, in $T$, from $x_0$ to $x$.  Note that each node $x\in V(T)$ is a $T$-ancestor and $T$-descendant of itself.  We say that a $T$-ancestor $a\in V(T)$ of $x\in V(T)$ is a \defin{strict} $T$-ancestor of $x$ if $a\neq x$.
The \defin{$T$-depth} of a node $x\in V(T)$ is the length of the path, in $T$, from $x_0$ to $x$.  For each node $x\in V(T)$, define
\[T_x := T[\{y\in V(T):\mbox{$x$ is a $T$-ancestor of $y$}\}] \]
to be the maximal subtree of $T$ rooted at $x$.

\note{PM}{I have significantly simplified this section, which should make the referees happy. I will read it very carefully again tomorrow}

We begin with a standard technique that allows us to work with a normalised tree-decomposition whose tree has same vertex set as the graph it decomposes:

@@

  Next, remove the node $x_0$ from $T_0$, and consider any node $a\in V(T_0)$ such that there is no vertex $x\in V(H)$ with $f(x)=a$.  In this case, $B_{a}\subseteq B_{a'}$ where $a'$ is the parent of $a$ since any $x\in B_a\setminus B_{a'}$ would have $f(x)=a$.  In this case, contract the edge $aa'$ in $T_0$, eliminating the node $a$.  Repeating this operation will eventually produce a width-$t$ tree-decomposition of $(B_x:x\in V(T_0))$ where $f$ is a bijection between $V(H)$ and $V(T_0)$.  Renaming each node $a\in V(T_0)$ as $f^{-1}(a)$ gives a tree-decomposition $(B_x:x\in V(T))$ with $V(T)=V(H)$.  By the definition of $f$, the tree-decomposition $(B_x:x\in V(T))$ satisfies \tref{subtree-root}.
  % To see that $(B_x:x\in V(T))$ satisifies \tref{ancestor-edge}, observe that,
  % if $xy\in E(H)$, then at least one of $x$ or $y$ is contained in $B_z$ for every node $z$ on the path from $x$ to $y$ in $T$.  If neither $x$ nor $y$ is an ancestor of the other, then some node $z$ on this path has $T$-depth less than that of $x$ and $y$.  If $x\in B_z$ this contradicts the fact that $x$ is the root of $T[x]$.  If $y\in B_z$ this contradicts the fact that $y$ is the root of $T[y]$.
\end{proof}

@@

%\subsection{Generalized Tripod Partitions}
% 
% 
% \begin{figure}[htbp]
%   \begin{center}
%     \includegraphics{figs/tripoddo}
%   \end{center}
%   \caption{The sets $Y_x$, $F_x$, and $V_x$ associated with $x\in V(T)$
%   and the ancestors $a_1,\ldots,a_{t'}$ of $X$ such that $F_x \subseteq \bigcup_{i=1}^{t'} Y_{a_i}$.}
%   \label{fig:generalized-tripod}
% \end{figure}

% The following lemma shows how to interpret an $H$-partition of $G$ and a tree-decomposition of $H$ as a `hierarchical' decomposition of $G$; refer to \cref{fig:generalized-tripod}.

% For a vertex $v$ of $G$, $N_G(v):=\{w\in V(G):vw\in E(G)\}$ and $N_G[v]:=N_G(v)\cup\{v\}$.  For a vertex set $S\subseteq V$, $N_G(S):=\bigcup_{v\in S} N_G(S)\setminus S$ and $N_G[S]:=N_G(S)\cup S$.

% \note{PM}{
% Guide to old notation:
% \begin{compactitem}
%   \item $V_x := \bigcup_{y\in V(T_x)} Y_y = G_x$
%   \item $F_x:=\{w\in V(G): vw\in E(G), v\in V_x,\, w\not\in V_x\}= N_G(G_x)$
%   \item $N_x:=V_x\cup F_x=N_G[G_x]$
% \end{compactitem}
% }
% 
% \begin{lem}\label{generalized-tripod}
%   Let $G$ be a graph; let $\mathcal{L}:=\langle L_1,\ldots,L_h\rangle$ be a layering of $G$; let $\mathcal{Y}:Y_x: x\in V(H))$ be an $H$-partition of $G$ of layered width at most $\ell$ with respect to $\mathcal{L}$ where $H$ has treewidth at most $t$; and let $\mathcal{T}:=(B_x:x\in V(T))$ be a tree-decomposition of $H$ satisfying the conditions of \cref{nice-decomposition}. 
%   For each $x\in V(T)$, let $H_x:=V(T_x)$ and let $G_x:=\bigcup_{y\in H_x}Y_y$.
% 
%   \begin{compactenum}[(Y1)]
%     % \item\ylabel{thickness} $\mathcal{Y}=(Y_x: x\in V(T))$ is a partition of $V(G)$ of layered width at most $\ell$ with respect to $\mathcal{L}$.
%     % \item\ylabel{separator} For each $x\in V(T)$, $N_G(\phi(T_x))$ separates $\phi(T_x)$ from $N
%     % there is no edge $vw\in E(G)$ with $v\in V_x$ and $w\in V(G)\setminus N_x$.
% 
%     \item\ylabel{ancestor-edge} For each $x\in V(T)$, there is a set $\{a_1,\ldots,a_{t'}\}$ of $t'\le t$ strict $T$-ancestors of $x$ such that $N_G(G_x) \subseteq \bigcup_{i=1}^{t'} Y_{a_i}$.
%   \end{compactenum}
% \end{lem}
% 
% \begin{proof}
%   Let $\{a_1,\ldots,a_{t'}\}:= B_x\setminus\{x\}$.  Then, by the definition of a normalized tree decomposition $N_H(H_x))\subseteq\{a_1,\ldots,a_{t'}\}$.  Therefore $N_G(G_x)\subseteq \bigcup_{i=1}^{t'} Y_{a_i}$.
% \end{proof}

% \referee{1}{1. Lemma 3: This lemma is really about ``normalised'' tree-decompositions, and says that for each $x \in V(T)$, the set $V(T_x)$ has at most $t$ neighbours in $H$. Call this property (T3) and remove (Y1) through (Y5), which should now be obvious.}
% 
% \note{PM}{I've added to the \texttt{response.tex} file explaining some of this.  Each of (Y1), (Y3), (Y4), and (Y5) have a one-sentence justification. The justification for (Y2) is just definition unpacking and then using the observation that the referee recommends.}
% 
% \note{PM}{I plan to do an experiment with the referee's suggested notation, in a separate git branch to see how it turns out.}
% 
% Before proving \cref{generalized-tripod} we point out more properties that are immediately implied by it:
% 
% \begin{compactenum}[(Y1)]\setcounter{enumi}{2}
%   \item\ylabel{y-subsets} $Y_x\subseteq V_x$ for every $x\in V(T)$. \note{PM}{$Y_x$, for $x\in V(T)$ is not defined!}
%   \item\ylabel{containment-i} $V_x\subseteq V_a$ for every $T$-ancestor $a$ of $x$.
%   \item\ylabel{containment-ii}$N_x\subseteq N_a$ for every $T$-ancestor $a$ of $x$.
% \end{compactenum}

% Property~\yref{y-subsets} follows from the fact that $V_x$ is the union of several sets, one of which is $Y_x$.  Property~\yref{containment-i} follows from the definition of $V_x$ and the fact that $V(T_x)\subseteq V(T_a)$. To show Property~\yref{containment-ii} first note that, by \yref{containment-i} it suffices to consider vertices $w\in F_x=N_x\setminus V_x$. By definition, every vertex $w\in F_x$ is adjacent, in $G$, to a vertex $v\in V_x$.  By \yref{containment-i}, $v\in V_a$, so $w$ is either in $V_a$ or $w$ satisfies the condition $vw\in E(G)$, $v\in V_a$, and $w\not\in V_a$, so $w\in F_a$.  In either case $w\in N_a=V_a\cup F_a$.  Note that none of \yref{y-subsets}--\yref{containment-ii} depends on \yref{ancestor-edge} (which is important, since \yref{containment-i} is used to establish \yref{ancestor-edge} in the following proof).
% 
% 
% \begin{proof}[Proof of \cref{generalized-tripod}]
%   % Property~\yref{thickness} follows immediately from the fact that $V(T)=V(H)$  and the fact that $\mathcal{Y}$ has layered width at most $\ell$ with respect to $\mathcal{L}$.
%   Property \yref{separator} is immediate from the definitions of $F_x$ and $N_x$.  In particular, $(N_x,V(G)\setminus V_x)$ is a separation of $G$ with $F_x=N_x\cap(V(G)\setminus V_x)$.
% 
%   To establish Property~\yref{ancestor-edge}, consider some vertex $w\in F_x$.  Since $w\in F_x$, there exists an edge $vw\in E(G)$ with $v\in V_x$ and $w\not\in V_x$.  Since $v\in V_x$, $v\in Y_{x'}$ for some $T$-descendant $x'$ of $x$ (possibly $x=x'$). Since $\mathcal{Y}$ is a partition, $w\in Y_{a}$ for some $a\not\in V(T_x)$.  Since $vw\in E(G)$, we have $x'a\in E(H)$.  By \tref{ancestor-edge}, one of $a$ or $x'$ is a $T$-ancestor of the other. Since $w\in Y_a\subseteq V_a$ and $w\not\in V_x\supseteq V_{x'}$, \yref{containment-i} rules out the possibility that $x'$ is a $T$-ancestor of $a$. Therefore, $a$ is a $T$-ancestor of $x$ which is a $T$-ancestor of $x'$.  Let $z_0,\ldots,z_r$ be the path in $T$ from $z_0:=x'$ to $z_r:=a$.  Since $x'a\in E(H)$, at least one of $a$ or $x'$ is in $B_{z_i}$, for each $i\in\{0,\ldots,r\}$ and there is at least one $i$ such that $B_{z_i}$ contains both $x'$ and $a$.  However, by \tref{subtree-root} $x'$ is not contained in $B_{z_i}$ for any $i\in\{1,\ldots,r\}$.  Therefore $a\in B_{z_i}$ for each $i\in\{0,\ldots,r\}$.  In particular, $a$ is contained in $B_x$.
%   Property~\yref{ancestor-edge} now follows from the fact that $|B_x|\le t+1$ and $B_x$ contains $x$.
% \end{proof}

% \referee{2}{Page 7.  In the very last sentence, I think $B_{x_i}$ should be $B_{z_i}$
% (twice).  Moreover, the way the proof is written, it seems to suggest
% that we only get that $a \in B_{z_i}$ for each $i \in \{1, \dots, r\}$.
% The conclusion that $a \in B_{z_i}$ for each $i \in \{0, \dots, r\}$ is
% true though.  In a normalized tree-decomposition, if $xy$ is an edge and
% $x$ is a $T$-ancestor of $y$, then $x$ must be in $B_y$ (since the subtrees $T_x$
% and $T_y$ intersect).}

%\note{PM}{I addressed the preceding referee's.}

@@

% \referee{1}{2. Proof of Theorem 9: There is too much notation throughout. As
% 	one example, it is hard to remember which of $T_x$, $Y_x$, $V_x$, $F_x$, $N_x$,
% 	$S_x$, $X_v$, $B_x$, $C_x$ is which. If $\phi(S)$ is written for the subset of $V(G)$
% 	corresponding to a set $S \subseteq V(H)$ and appropriate shorthand and
% 	standard notation for neighbourhoods are used, then $Y_x$, $V_x$, $F_x$, and
% 	$N_x$ can be replaced by $\phi(x)$, $\phi(T_x)$, $N(\phi(T_x))$, and $V(G) \setminus N[\phi(T_x)]$.
% 	The set $X_v$ is only used in the statement of Claim~1, so it does not
% 	need a name; say what these vertices are explicitly. It may be more
% 	indicative to write $\phi'(x)$ for $S_x$ and $B'_x$ for $C_x$.}

\begin{proof}
Let $\mathcal{L}:=\langle L_1,\ldots,L_h\rangle$ be a layering of $G$; let $\mathcal{Y}:=(Y_x: x\in V(H))$ be an $H$-partition of $G$ of layered width at most $\ell$ with respect to $\mathcal{L}$; and let $\mathcal{T}:=(B_x:x\in V(T))$ be a tree-decomposition of $H$ satisfying the conditions of \cref{nice-decomposition}.   
% For each $x\in V(T)$, let $H_x:=V(T_x)$ and let $G_x:=\bigcup_{y\in H_x}Y_y$.

% Apply \cref{generalized-tripod} to $G$ and let $\mathcal{L}$, $\mathcal{Y}$, $\mathcal{T}$, $T$, $H_x$, and $G_x$ be defined as in \cref{generalized-tripod}, where the partition $\mathcal{Y}$ has width $\ell$ with respect to the layering $\mathcal{L}$.

For a node $x\in V(T)$, we say that a shortcut $P\in\SS$ \defin{crosses} $x$ if $Y_x$ contains an internal vertex of $P$.  In other words, if $P=v_0,\ldots,v_r$ and $\{v_1,\ldots,v_{r-1}\}\cap Y_x\neq\emptyset$.  We say that a vertex $v\in V(G)$ \defin{participates} in $x$ if $v\in Y_x$, or $\SS$ contains a shortcut $P$ with $v\in V(P)$ and $P$ crosses $x$.  For each $v\in V(G)$, let $a(v)$ denote the lowest common $T$-ancestor of all the nodes $x\in V(T)$ in which $v$ participates.

% \begin{clm}
%   For each $v\in V(G)$, $v$ participates in $a(v)$.
% \end{clm}
% 
% \begin{proof}
%   Let $Z := \{v\} \cup \{\{v_1,\ldots,v_{r-1}\}:(v_0,\ldots,v_r)\in\SS, v\in \{v_0,\ldots,v_r\}\}$. Then $G[Z]$ is connected because $Z$ is the union of (vertex sets of) paths in $G$, each of which contains $v$.
% 
%   We claim that $v$ participates in a node $x\in V(T)$ if and only if $Z\cap Y_x\neq\emptyset$.  If $v$ participates in $x$ then either $v\in Y_x$, so $Z\cap Y_x\supseteq\{v\}$; or $v\in \{v_0,\ldots,v_r\}$ for some shortcut $(v_0,\ldots,v_r)\in\SS$ that crosses $x$, so $Z\cap Y_x\supseteq \{v_i\}$ for some $i\in\{1,\ldots,r-1\}$.  In the other direction, if $Z\cap Y_x\neq\emptyset$, then either $Z\cap Y_x\supseteq \{v\}$, so $v\in Y_x$; or $Z\cap Y_x\supseteq \{v_i\}$ where $i\in\{1,\ldots,r\}$, $(v_0,\ldots,v_r)\in\SS$ and $v\in\{v_0,\ldots,v_r\}$, so $v\in V(P)$ for a path $P=(v_0,\ldots,v_r)\in\SS$ that crosses $x$.
% 
%   Let $X_H:=\{x\in V(H): Z\cap Y_x\neq\emptyset\}$.  Therefore $X_H$ contains every every node in which $v$ participates.  The connectivity of $G[Z]$ implies that $H[X_H]$ is connected. The fact that $\mathcal{T}$ is a normalized tree decomposition of $H$ implies that some node $a\in X_H$ is a $T$-ancestor of every node in $X_H$.  Thus, $v$ participates in the node $a=a(v)$ that is the lowest common $T$-ancestor of every node in which $v$ participates, as required.  
% \end{proof}

% \begin{proof}
%   Let $a(v)$.  If $v\in Y_{a(v)}$ then $v$ participates in $a(v)$, by definition, and we are done.  Otherwise, there are two nodes $x,y\in V(T)$ in which $x$ participates and whose lowest common $T$-ancestor is $a(v)$.
% 
%   There are two cases to consider:  
%   \begin{compactenum}
%     \item If $v\in Y_x$ then $v\not\in Y_y$.  Since $v$ participates in $Y_y$, there exist a shortcut $P:=v_0,\ldots,v_r$ in $\SS$ with $v=v_\alpha$ and $v_\beta\in Y_y$ for some $\alpha\in\{0,\ldots,r\}$ and $\beta\in\{1,\ldots,r-1\}$.  Without loss of generality, $\alpha < \beta$.
% 
%     For each $i\in\{0,\ldots,r\}$, let $Y_{x_i}$ be the part of $\mathcal{Y}$ that contains $v_i$.  Since $G[\{v_\alpha,\ldots,v_\beta\}]$ is connected, $H[\{x_\alpha,\ldots,x_\beta\}]$ connected.  Since $\mathcal{T}$ is a normalized tree decomposition of $H$, $x_j$ is a $T$-ancestor of every node in $x_\alpha,\ldots,x_\beta$ for some $j\in\{\alpha,\ldots,\beta\}$.      In particular $x_j$ is a $T$-ancestor of $x=x_\alpha$ and $y=x_\beta$, so $x_j$ is a $T$-ancestor of $a(v)$.  Since $v_j\in V(P)$ and $v_j\in Y_{x_j}$ this implies that $v$ participates in $x_j$.  Therefore $v$ participates in a $T$-ancestor of $a(v)$, which can only mean that $v$ participates in $a(v)$.
% 
%     \item If $v\not\in Y_x$ and $v\not\in Y_y$ then there are shortcuts $P_1,P_2\in\SS$ with $v\in V(P_1)$, $v\in V(P_2)$ where $P_1$ has an internal vertex $w_1\in Y_x$ and $P_2$ has an internal vertex $w_2\in Y_y$.    
% 
% 
%     .  Since $v$ participates in $Y_y$, there exist a shortcut $v_0,\ldots,v_r$ in $\SS$ with $v=v_\alpha$ and $v_\beta\in Y_y$ for some $\alpha\in\{0,\ldots,r\}$ and $\beta\in\{1,\ldots,r-1\}$.  Without loss of generality, $\alpha < \beta$.
% 
%     For each $i\in\{0,\ldots,r\}$, let $Y_{x_i}$ be the part of $\mathcal{Y}$ that contains $v_i$.  Since $G[\{v_\alpha,\ldots,v_\beta\}]$ is connected, $H[\{x_\alpha,\ldots,x_\beta\}]$ connected.  Since $\mathcal{T}$ is a normalized tree decomposition of $H$, $x_j$ is a $T$-ancestor of every node in $x_\alpha,\ldots,x_\beta$ for some $j\in\{\alpha,\ldots,\beta\}$.  In particular $x_j$ is a $T$-ancestor of $x=x_\alpha$ and $y=x_\beta$, so $x_j$ is a $T$-ancestor of $a(v)$. Since $v_j\in Y_{x_j}$, and $v\not\in Y_{a(v)}$ this implies that $v$ participates in $Y_{x_j}$.  Therefore $v$ participates in a $T$-ancestor of $a(v)$, which means that $v$ participates in $a(v)$.
% 
% 
%   \end{compactenum}
% \end{proof}

% We let $X_v$ denote the set of nodes $x\in V(T)$ such that $v$ participates in $x$.

% \begin{clm}\label{x-v-ancestor}
%   For any $v\in V(G)$ there exists a (unique) node $a(v)\in X_v$ such that
%   $a(v)$ is a $T$-ancestor of every node in $X_v$.
% \end{clm}
% 
% \begin{proof}
%   Let $Z := \{v\} \cup \{\{v_1,\ldots,v_{r-1}\}:(v_0,\ldots,v_r)\in\SS, v\in \{v_0,\ldots,v_r\}\}$. Then $G[Z]$ is connected because $Z$ is the union of (vertex sets of) paths in $G$, each of which contains $v$.
% 
%   We claim that $v$ participates in a node $x\in V(T)$ if and only if $Z\cap Y_x\neq\emptyset$.  If $v$ participates in $x$ then either $v\in Y_x$, so $Z\cap Y_x\supseteq\{v\}$; or $v\in \{v_0,\ldots,v_r\}$ for some shortcut $(v_0,\ldots,v_r)\in\SS$ that crosses $x$, so $Z\cap Y_x\supseteq \{v_i\}$ for some $i\in\{1,\ldots,r-1\}$.  In the other direction, if $Z\cap Y_x\neq\emptyset$, then either $Z\cap Y_x\supseteq \{v\}$, so $v\in Y_x$; or $Z\cap Y_x\supseteq \{v_i\}$ where $i\in\{1,\ldots,r\}$, $(v_0,\ldots,v_r)\in\SS$ and $v\in\{v_0,\ldots,v_r\}$, so $v\in V(P)$ for a path $P=(v_0,\ldots,v_r)\in\SS$ that crosses $x$.
% 
% 
%   Let $X_H:=\{x\in V(H): Z\cap Y_x\neq\emptyset\}$.  The connectivity of $G[Z]$ implies that $H[X_H]$ is connected.  This implies that $T':=T[]
% 
% 
% 
%   %Property~\tref{ancestor-edge} implies that, for every $xy\in E(H)$, one of $x$ or $y$ is a $T$-ancestor of the other.
%   Choose $a(v)\in X_H$ to be the member of $X_H$ that does not have a strict $T$-ancestor in $X_H$.  Transitivity of the $T$-ancestor relationship, \tref{ancestor-edge}, and connectivity of $H[X_H]$ implies that such an $a(v)$ exists and is a $T$-ancestor of every node $x\in X_H$, as required.
% \end{proof}

% \note{DW}{Use $\SS$ for the given shortcut system and $\PP$ for the partition?}
% \note{PM}{Yes, I've done this.  Again, I was carefuly, but need to reread.}
